QueryMed: An Intuitive SPARQL Query Builder for Biomedical RDF Data

### 2010-02-15 trunk/docs/www2010/paper.tex
final submission

QueryMed is flexible enough to allow queries relevant to a wide range of biomedical topics, runs queries across multiple SPARQL endpoints, and is designed to be accessible to users who do not know the structure of the underlying ontologies used in describing the datasets, or the SPARQL query language to query the data. The system allows users to select the data sources that they wish to use, drawing on their specialized domain knowledge to decide the most appropriate data sources to query.  Users can add additional data sources if they are interested in querying endpoints that are not in the default list. After retrieval of the initial result set, query results can be filtered to improve their relevance.  The system also allows the user to exploit the underlying structure of the RDF data to improve query results. 
\end{abstract}

@@

\keywords{Biomedical Ontologies, SPARQL, Query Federation, Query Building, Semantic Web, User Interfaces}

@@

The quantity of publicly available data in the biomedical domain has dramatically increased over recent years.   Publicly available biomedical resources include data on drug discovery~\cite{Sharp, Goble}, clinical trials, diseases, disease genes, and phenotypes.  With the linked open data movement, the semantic web community has been very proactive in converting these rich information resources to RDF~\cite{LinkingData}.  In fact, the biomedical domain is among the early successes of the semantic web due to the rapidity with which the community has made its data available in RDF triple stores \cite{Yip}.

@@

To allow end users to exploit the abundance of useful biomedical data that is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and that also allow users to run federated queries on multiple SPARQL endpoints.  There is also a  need for efficient hybrid interfaces that allow both browsing and querying~\cite{Jentzsch}, since many currently available systems are linked data browsers such as the Tabulator~\cite{Tabulator}, which allow a user to navigate the data in an exploratory manner but lack support for filtering and querying the data in a user friendly manner.

@@

Answering many medically and biologically relevant questions requires searching, filtering, and combining information from multiple endpoints.  For example, a physician may know her patient's personal information, symptoms, current medications, and genotype. She may wish to determine the patient's treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--``based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?"--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and GO \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.

@@

Since the physician must search many databases in order to find an answer to her single question, she requires a system that can automatically run queries over multiple data sources. Also, the physician may not know SPARQL query syntax, the location of the SPARQL endpoints, or the structure of the relevant ontologies. She is likely to want an intuitive way to query and to display the query result.  Developing intuitive ways to query multiple data sources and display results is both an important and a challenging problem.  Our system, QueryMed, allows users with no knowledge of the SPARQL query language or the structure of the underlying ontologies to easily run queries across multiple SPARQL endpoints.

@@

This paper is organized as follows: Section \ref{background} provides background information on the semantic web and its relevance for modeling data in the biomedical domain. Section \ref{system} describes our system. Section \ref{related} discusses related work and illustrates how QueryMed differs from previous systems. Section \ref{future-work} outlines future work giving an outlook for Querymed. Finally section \ref{conclusion} summarizes the contributions of our system.

@@

Many major biological and biomedical data resources, including Gene Ontology, DailyMed, LinkedCT, and Diseasome, are currently available as RDF triplestores.   Almost all of these data sources are interlinked with each other. Integrating biomedical data across multiple data sources and automatically extracting specific  knowledge from web resources are crucial tasks for physicians and biologists. These semantic web resources represent valuable repositories of information that can be automatically mined for applications that require biological knowledge.

@@

Although many valuable resources in the biomedical domain are available in RDF, there are a number of challenges that must be addressed in order to make such resources accessible to physicians, patients, and life scientists.  One challenge is constructing systems that allow end users to run intuitive queries on biomedical data.  Users of biomedical resources are likely to have extensive domain knowledge, but be unfamiliar with the SPARQL query language syntax and with the structure of biomedical ontologies.  Almost all the SPARQL endpoints available today are offering a generic query interface that requires users to write SPARQL queries.  Also, no hints about the structure of data available from the SPARQL endpoint is given as the user is constructing the SPARQL query.  The user has to essentially construct the queries that are needed for her purpose by trial and error.  This can be a daunting task  especially for someone who is new to semantic web technologies.  Therefore, it is important to design user-friendly systems that allow users to take advantage of the wealth of structured biological knowledge available on the semantic web.  Another central challenge is designing systems that permit users to query multiple data sources simultaneously, since relevant biological data are often distributed among many sources \cite{Pasquier}.

@@

The QueryMed system allows users to easily query multiple biomedical data sources.  Queries can be run against a default list of datasources, or against a set of user-selected SPARQL endpoint.  The end user can easily input additional endpoints in order to utilize resources that are not included in the default list.  The system automatically translates the user input into a SPARQL query for each individual endpoint, combines the results, and returns them to the user.  The user can choose to refine the query by iteratively modifying the original query terms, and by filtering the result list.  The advanced query functionality of the QueryMed system allows the user to easily construct complex logical SPARQL queries that take advantage of the underlying structure of the ontologies. The simple user interface of the QueryMed system is designed to be intuitive for users with no knowledge of the SPARQL query language.

@@

A general overview of QueryMed architecture is shown in Figure \ref{fig:arch_overview}. The main components of the system are the user interface and the proxy server that takes input from the user and retrieves relevant biomedical data from remote SPARQL endpoints.   After a user submits a query from the user interface, the query is translated by the proxy server into individual SPARQL queries for each remote endpoint.  The query results are returned from the remote endpoints, combined by the proxy server, and presented to the user.  A detailed illustration of the parts of the QueryMed system is shown in Figure \ref{fig:arch_details}.

@@

The QueryMed user interface is designed to be intuitive for the end user, yet flexible enough to permit a broad range of interesting queries.  The basic query interface allows the user to run simple queries, and is designed for maximal ease of use.  The advanced search capabilities allow the user to easily construct complex logical queries that take advantage of the underlying structure of the biomedical ontologies.  The user interface also allows the user to iteratively refine queries, and displays the query results, which have been retrieved from multiple SPARQL endpoints.  The main components of the user interface are as follows:

@@

\item \textbf{Basic Query Interface:}  In order to make QueryMed easy to use, but also provide a flexible system that is capable of performing a broad variety of biomedically relevant queries, we provide an uncluttered basic search interface with only a  blank search box and two buttons as shown in Figure \ref{fig:initial_ui}. The ``Query All" button will perform a keyword-based query over pre-selected data sources. The advanced search option ``Refine Query" allows the user to view additional query options that are handled by the Advanced Query Interface.

@@

 For example, a physician interested in finding semantic web resources related to coronary artery disease could use “coronary artery disease”  as a search keyword in the input search box and select the "Query All" option.   The simplicity of the basic query interface allows the physician to rapidly and easily browse semantic web resources of interest.

@@

\item \textbf{Result Browser:} The user can view his query results organized by source in the Result Browser. The results are presented in a table with pagination.  The user can choose the number of results viewed at a time, search the columns based on some text value and also sort the columns as shown in Figure \ref{fig:results}.  This allows the user to perform additional filtering on the query results to display the most relevant results, and is particularly useful for refining queries that return a large number of results.

@@

For example, after searching for "coronary artery disease" in the basic query interface, the physician will see displayed in the Result Browser a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery 
disease. She can then filter the results using additional search terms. For example, she knows that 
the route of administration of the drug that she is interested in is injection, so she filters the drug query results on the route of administration field using the query term “injection.”   The additional filtering capabilities of the Result Browser allows her  to either browse a large number of query results, or refine her search to view only the most relevant query results.

\item \textbf{Advanced Query Interface:} The Advanced Query Interface allows the user to add new data sources, and to construct a complex SPARQL query that takes advantage of the structure of the underlying biomedical ontologies.  Thus, the end user can construct a targeted query without previous knowledge of the SPARQL language or the structure of the relevant biomedical ontologies.

When the ``Refine Query" in the Basic Query Interface is invoked, the user is provided with the default list of data sources in the system as shown in Figure \ref{fig:select}.  The user may select from this list relevant trusted data sources that they wish to query.  The user can also dynamically add additional data sources as shown in Figure \ref{fig:add_source}, allowing users to query endpoints of interest that are not included in the default list.

A physician who is interested in using the QueryMed system to find relevant clinical trials for 
her patient can use the Advanced Query Interface to add additional relevant datasources .  For example, she might want to use the clinical trial database LinkedCt, which is 
not in the default set of endpoints.  So, she selects the “Refine 
Query” option to select the endpoints to search, and then the “Add” option 
to include an additional endpoint. After entering the name 
and URL of the LinkedCT endpoint, she can able to search 
for clinical trials for which her patient may be eligible.

@@

\begin{figure}
\centering
\epsfig{file=images/add_source.jpg,  width=3in}
\caption{Add New Data Source}
\label{fig:add_source}
\end{figure}

When a data source is selected, the properties list is automatically populated with all distinct properties available at the selected endpoint. If the user wants to restrict her query based on one or more properties available, she can specify values for the desired properties.  Because the QueryMed system automatically displays a list of properties to the user, users can improve their queries using the underlying structure of the data to obtain more relevant query results. 
%The properties list is generated by running a SPARQL query of the form given in Figure \ref{fig:SPARQL_props} at the specified data source.

@@

The user can choose to perform exact queries, or filter results on specific keywords. If the user does not know the specific value for a property, she can instead specify a keywords using the “FILTER” option. She also can choose logical operators to connect the various parts of their query. When “AND” is used it will be appended as a basic triple pattern to the SPARQL query, i.e. as a conjunction. When “OR” is used, the specified graph pattern is made to disjunct with the rest of the query with the SPARQL UNION operator. If no logical operator is specified, “AND” will be used by default.  The advanced query feature is capable of dynamically constructing complex SPARQL queries, such as the query shown in Figure \ref{fig:SPARQL_adv}.

@@

For example, the physician might want to further refine her search for cardiovascular diseases. She 
uses the advanced search option to construct a boolean query 
that takes advantage of the underlying structure of the RDF 
data in the database. She searches Diseasome for a list of 
diseases whose class is “Cardiovascular” or for which the associated gene is “ABCA1.” Using the QueryMed advanced 
search interface, the complex SPARQL query corresponding 
to her question is automatically constructed (Figure \ref{fig:SPARQL_adv}), and she can 
view the query results conveniently displayed in the Results Browser.

\begin{figure*}
\centering
\epsfig{file=images/add_props.jpg, width=6in}
\caption{The Advanced Search feature allows the user to perform exact or pattern-matching queries connected by user-specified logical operators over specific properties in given resources, taking advantage of the structure in the RDF data.}
\label{fig:add_props}
\end{figure*}

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT distinct ?disease WHERE { 
{?x <http://www.w3.org/2000/01/rdf-schema#label> 
?disease 
FILTER regex(?disease, 
"coronary artery disease", "i"). 
?x <http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseasome/class> 
<http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseaseClass/Cardiovascular>} 
UNION { ?x <http://www4.wiwiss.fu-berlin.de/
diseasome/resource/diseasome/associatedGene>
 <http://www4.wiwiss.fu-berlin.de/diseasome/
 resource/genes/ABCA1>.}
}
\end{Verbatim}
\caption{A complex SPARQL query that takes advantage of the underlying data structure can be dynamically constructed using the advanced query feature of the QueryMed system.}
\label{fig:SPARQL_adv}
\end{figure}

@@

\begin{figure}
\centering
\epsfig{file=images/facet_creator.jpg, width=3in}
\caption{Dynamic Facet Creator and Filter}
\label{fig:facet_creator}
\end{figure}

\begin{figure}
\centering
\epsfig{file=images/results.jpg, width=3.5in}
\caption{A sample result table}
\label{fig:results}
\end{figure}

@@

\item \textbf{Source Manager:} The Source Manager reads the source list and populates the default query list on the user interface. It also keeps track of the default endpoints, the currently selected endpoints, and the endpoints that have been dynamically added.

@@

% The user has to specify the data source(s) to query, and can give specific values for one or many of the properties available for the data available to restrict the query space. 
\item \textbf{Additional Services:} In order to make the user interface more user friendly, the following two services perform some ``behind the scenes" work.

\begin{enumerate}
\item \textbf{Get Properties:} This service takes as input an individual source, and returns an array of all properties for that source.
\item \textbf{Get Property Values:} This service takes as input an individual source and a specific property.  It returns a list of the possible values that this property can take.
\end{enumerate}

@@

\item The proxy server enables us to avoid cross domain XML-HTTP-Request errors in accessing SPARQL endpoints in various domains.

@@

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT ?projection_1 ?projection_2 ... 
	?projection_n
WHERE {?x source:property ? projection 
	FILTER regex(? projection, '" + 
	input +"', 'i)
}
... //Other property filers are to follow
\end{Verbatim}
\caption{Minimal SPARQL Query Structure in the ``Query All" Case}
\label{fig:min_query}
\end{figure}

@@

When the user runs an advanced query, the proxy server takes as input the data sources to be queried, the properties to be queried, the values for each of the selected properties, and the relationship between the property-value pairs and the original user-specified query term. This information is passed to the proxy server as a JSON object.

@@

The basic query interface is designed to be as simple as possible, so that users who have little previous experience running queries on SPARQL endpoints can easily and rapidly find query results.  The advanced query functionality of our system allows users the flexibility to construct complex SPARQL queries.  The QueryMed system allows the user to construct queries that take advantage of the underlying structure of the data exposed by the individual data endpoints, without requiring previous knowledge of this structure.

\subsubsection{Implementation}

@@

For all selected endpoints, without local data (i.e. no cache on the proxy server), the property values took longer to load.  The difference in running time between \emph{with local data} and \emph{without local data} is approximately three orders of magnitude.  The network latency and the slow performance of the remote SPARQL endpoints as compared to the local data cache can be accounted for the increase in running time for the queries in the two cases.  We also noted that the running time of the subsequent iterations of the same query is significantly less than the initial query, probably due to browser caching. When comparing the query execution time amongst the data sources, we see that the Drugbank takes the longest, probably due to the greater size of the Drugbank data as compared to the other three data sources~\footnote{As of February 15\textsuperscript{th} 2010, Diseasome contains 91,182 triples, DailyMed contains 164,276 triples, Sider contains 192,515 triples, and DrugBank contains 765,936 triples) \cite{Diseasome, DailyMed, Sider, Drugbank}}.

@@

Therefore, it is fair to say that bottleneck in the running time of our system is running queries that must retrieve many triples from slow remote SPARQL endpoints.  By managing our own cache to contain the data most likely to be needed on the proxy server, we were able to dramatically decrease running time.

@@

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT DISTINCT ?property
WHERE { [] ?property [] }
ORDER BY ?property
\end{Verbatim}
\caption{SPARQL query to retrieve all Properties. This was used to measure the performance of implementing a local cache.}
\label{fig:SPARQL_props}
\end{figure}

\begin{figure}
\centering
\epsfig{file=images/chart.jpg, width=3in}
\caption{Comparison of Query Execution Times (measured in milliseconds) at Various SPARQL Endpoints With and Without Local Data (Cache) (average of three trials)}
\label{fig:chart}
\end{figure}

@@

\begin{table*}
\begin{center}
\begin{tabular}{| p{1.8cm} | p{2cm} | p{1.8cm} | p{1.8cm}| p{1.8cm}| p{1.8cm}| p{1.8cm}| p{1.8cm} |}
\hline
 & \textbf{Hybrid Interface (Combines Querying \& Browsing)} & \textbf{Provides Local Caching} & \textbf{Queries Multiple Sources} & \textbf{Dynamic Addition of Sources} & \textbf{Allows Keyword Queries} & \textbf{Open Source} & \textbf{GUI} \\
\hline
\textbf{QueryMed}  & Yes & Yes & Yes & Yes & Yes & Yes & Yes  \\
\hline
\textbf{SMART}  & No & Yes & Yes & No & No & Yes & Yes  \\
\hline
\textbf{DARQ}  & No &  No & Yes & N/A & No & Yes & No  \\
\hline
\textbf{GoWeb}   & Yes & No  & Yes & No & Yes & No & Yes  \\
\hline
\textbf{BioGateway}   & Yes & Yes & Yes & No & No & Yes & Yes  \\
\hline
\textbf{Twinkle}   & No & No & Yes & No & No & Yes & Yes  \\
\hline
\end{tabular}
\caption{Comparison of selected features of the QueryMed system with other related systems.}
\label{comparison}
\end{center}
\end{table*}

A number of existing tools aim to provide a user-friendly interface for browsing semantic web data, or to allow users to perform federated queries.  Several of these are described below.

@@

The SMART query tool is a web-based application designed to allow biologists to run SPARQL queries over multiple endpoints.  Queries to the SMART system are constructed using a descriptive logic query written in the natural-language like Manchester OWL syntax \cite{Battista}.  Major differences between the QueryMed system and SMART include the ability to construct queries intuitively by specifying keywords for user selected properties, to iteratively refine query results, and to dynamically add additional endpoints in the QueryMed system.

@@

GoWeb and BioGateway are two additional systems designed for answering queries on biomedical data  \cite{Dietze, Antezana}.  GoWeb allows users to a hybrid search, running keyword-based queries and then filtering based on ontological concepts. However, while GoWeb functions as a search engine that incorporates ontological background knowledge to improve search results, QueryMed is a query translation system.  BioGateway provides a web interface to query a provided SPARQL endpoint that integrates multiple biomedical resources, but does not allow the user to run queries over multiple endpoints or to dynamically add endpoints.

@@

Most SPARQL query engines are designed to run queries against individual endpoints. But it is often useful to draw on multiple web resources in answering a query. There are a number of systems, including the DARQ \cite{Quilitz} and CALO query manager \cite{Ambite} systems, designed to allow the user to run integrated queries against multiple SPARQL endpoints.

>>>>

QueryMed is flexible enough to allow queries relevant to a wide range of biomedical topics, and runs queries across multiple SPARQL endpoints. It is designed to be accessible to users who are not familiar with the structure of the underlying ontologies used in describing the datasets, or with the SPARQL query language. The system allows users to select the data sources that they wish to use, drawing on their specialized domain knowledge to decide the most appropriate data sources to query.  Users can add additional data sources if they are interested in querying endpoints that are not in the default list. After retrieval of the initial result set, query results can be filtered to improve their relevance.  The system also allows the user to exploit the underlying structure of the RDF data to improve query results. 
\end{abstract}

@@

\keywords{Biomedical Ontologies, SPARQL Query Building, Query Federation, Semantic Web, User Interfaces}

@@

The quantity of publicly available data in the biomedical domain has dramatically increased over recent years.   Publicly available biomedical resources include data on drug discovery~\cite{Sharp, Goble}, clinical trials, diseases, disease genes, and phenotypes.  With the linked open data movement, the semantic web community has been very proactive in converting these rich information resources to RDF~\cite{LinkingData}.  In fact, the biomedical domain is among the early successes of the semantic web, due to the rapidity with which the community has made its data available in RDF triple stores \cite{Yip}.

@@

To allow end users to exploit the abundance of biomedical data that is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and that also allow users to run federated queries on multiple SPARQL endpoints.  There is also a  need for efficient hybrid interfaces that allow both browsing and querying data~\cite{Jentzsch}. Many currently available systems are linked data browsers such as the Tabulator~\cite{Tabulator} which permit users to navigate the data in an exploratory manner, but lack support for filtering and querying the data in a user friendly manner.

@@

Answering many medically and biologically relevant questions requires searching, filtering, and combining information from multiple data sources.  For example, a physician may know her patient's personal information, symptoms, current medications, and genotype. She may wish to determine the patient's treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--``based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?"--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and Gene Ontology \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.

@@

Since the physician must search many data sources in order to find an answer to her single question, she requires a system that can automatically run queries over multiple SPARQL endpoints. Also, the physician may not know SPARQL query syntax, the location of the SPARQL endpoints, or the structure of the relevant ontologies. She is likely to want an intuitive way to query and to display the query results.  Developing intuitive ways to query multiple SPARQL endpoints and to display results is both an important and a challenging problem.  Our system, QueryMed, allows users with no knowledge of the SPARQL query language or the structure of the underlying ontologies to easily run queries across multiple SPARQL endpoints.

@@

This paper is organized as follows:  Section \ref{background} provides background information on the semantic web and its relevance for modeling data in the biomedical domain. Section \ref{system} describes our system. Section \ref{related} discusses related work and illustrates how QueryMed differs from previous systems. Section \ref{future-work} outlines future work. Finally section \ref{conclusion} summarizes the contributions of our system.

\pagebreak

@@

Many major biological and biomedical data resources, including Gene Ontology, DailyMed, LinkedCT, and Diseasome, are currently available as RDF triplestores.   Almost all of these data sources are interlinked. Integrating biomedical data across multiple data sources and automatically extracting specific  knowledge from web resources are crucial tasks for physicians and biologists. These semantic web resources represent valuable repositories of information that can be automatically mined for applications that require biological knowledge.

@@

Although many valuable resources in the biomedical domain are available in RDF, there are a number of challenges that must be addressed in order to make such resources accessible to physicians, patients, and life scientists.  One challenge is constructing systems that allow end users to run intuitive queries on biomedical data.  Users of biomedical resources are likely to have extensive domain knowledge, but be unfamiliar with query language syntax and with the structure of biomedical ontologies.  Almost all the SPARQL endpoints available today offer generic query interfaces that require users to manually write their queries.  This can be a daunting task  especially for someone who is new to semantic web technologies.  Therefore, it is important to design user-friendly systems that allow users to take advantage of the wealth of structured biomedical knowledge available on the semantic web.  Another central challenge is designing systems that permit users to query multiple data sources simultaneously, since relevant biomedical data are often distributed among many sources \cite{Pasquier}.

@@

The QueryMed system allows the user to easily query multiple biomedical data sources.  Queries can be run against a default list of SPARQL endpoints, or against a set of user-selected endpoints.  The end user can easily input additional endpoints in order to utilize resources that are not included in the default list.  The system automatically translates the user input into a SPARQL query for each individual endpoint, executes the query, combines the results, and returns them to the user.  The user can choose to refine the query by iteratively modifying the original query terms, and by filtering the result set.  The advanced query functionality of the QueryMed system allows the user to easily construct complex logical SPARQL queries that take advantage of the underlying structure of the ontologies. The simple user interface of the QueryMed system is designed to be intuitive for users with no knowledge of the SPARQL query language.

@@

A general overview of QueryMed architecture is shown in Figure \ref{fig:arch_overview}. The main components of the system are the user interface and the proxy server.   After a user submits a query from the user interface, the query is translated by the proxy server into individual SPARQL queries for each remote endpoint.  The query results are returned from the remote endpoints, combined by the proxy server, and presented to the user.   Figure \ref{fig:arch_details} presents a detailed illustration of the parts of the QueryMed system.

@@

The QueryMed user interface is designed to be intuitive for the end user, yet flexible enough to permit a broad range of interesting queries.  The basic query interface allows the user to run simple queries, and is designed for maximal ease of use.  The advanced search capabilities enable the user to easily construct complex logical queries that take advantage of the underlying structure of the biomedical ontologies.  The user interface also allows the user to iteratively refine queries, and displays the query results, which have been retrieved from multiple SPARQL endpoints.

The main components of the user interface are as follows:

@@

\item \textbf{Basic Query Interface:}  In order to make QueryMed easy to use, but also provide a flexible system that is capable of performing a broad variety of biomedically relevant queries, we provide an uncluttered basic search interface as shown in Figure \ref{fig:initial_ui}. The ``Query All" button performs a keyword-based query over the default set of  data sources. The advanced search option ``Refine Query" allows the user to access the Advanced Query Interface.

@@

 For example, a physician interested in finding semantic web resources related to coronary artery disease could use ``coronary artery disease"  as a search keyword in the input search box and select the ``Query All" option.   This query searches all default data sources for resources related to coronary artery disease.  The simplicity of the basic query interface allows the physician to rapidly and easily browse semantic web resources of interest.

@@

\item \textbf{Result Browser:} The user can view her query results organized by source in the Result Browser. The results are presented in a table with pagination.  The user can choose the number of results viewed at a time, search the columns based on some text value and also sort the columns as shown in Figure \ref{fig:results}.  This allows the user to perform additional filtering on the query results to display the most relevant results.  This feature is particularly useful for refining queries that return a large number of results.

\begin{figure}
\centering
\epsfig{file=images/results.jpg, width=3.5in}
\caption{Sample Result Table}
\label{fig:results}
\end{figure}

@@

For example, after searching for ``coronary artery disease" in the basic query interface, the physician will see displayed in the Result Browser a list of disease names in the Diseasome database and drugs in DailyMed and Drugbank that relate to coronary artery 
disease. She can then filter the results using additional search terms. For example, she knows that 
the route of administration of the drug that she is interested in is injection, so she filters the drug query results on the route of administration field using the query term ``injection."   The additional filtering capabilities of the Result Browser allow her  either to browse a large number of query results, or refine her search to view only the most relevant results.

\item \textbf{Advanced Query Interface:} The Advanced Query Interface allows the user to add new data sources, and to construct a complex SPARQL query that takes advantage of the structure of the underlying biomedical ontologies.  Thus, the end user can construct a targeted query without previous knowledge of the SPARQL language or the structure of the relevant ontologies.

When the ``Refine Query" in the Basic Query Interface is invoked, the user is provided with the default list of data sources as shown in Figure \ref{fig:select}.  The user may select from this list relevant data sources that they wish to query.  The user can also dynamically add additional data sources as shown in Figure \ref{fig:add_source}, allowing her to query endpoints of interest that are not included in the default list.

A physician who is interested in using the QueryMed system to find relevant clinical trials for 
her patient can use the Advanced Query Interface to add additional relevant datasources.  For example, she might want to use the clinical trial database LinkedCT, which is 
not in the default set of endpoints.  So, she selects the “Refine 
Query” option to select the endpoints to search, and then the “Add” option 
to include an additional endpoint. After entering the name 
and URL of the LinkedCT endpoint, she can able to search 
for clinical trials for which her patient may be eligible.

@@

\begin{figure}
\centering
\epsfig{file=images/add_source.jpg,  width=3in}
\caption{The user can dynamically add additional SPARQL endpoints.}
\label{fig:add_source}
\end{figure}

When a data source is selected, the properties list is automatically populated with all distinct properties available at the selected endpoint. The user can specify values for relevant properties to restrict the search space.
%The properties list is generated by running a SPARQL query of the form given in Figure \ref{fig:SPARQL_props} at the specified data source.

@@

The user can choose to perform exact queries, or filter results on specific keywords. If the user does not know the specific value for a property, she can instead specify a keywords using the “FILTER” option. She also can choose logical operators to connect the various parts of her query. When “AND” is used it will be appended as a basic triple pattern to the SPARQL query, i.e. as a conjunction. When “OR” is used, the specified graph pattern is made to disjunct with the rest of the query with the SPARQL UNION operator. If no logical operator is specified, “AND” will be used by default.  The advanced query feature is capable of dynamically constructing complex SPARQL queries, such as the query shown in Figure \ref{fig:SPARQL_adv}.

@@

For example, the physician might want to further refine her search for cardiovascular diseases. She 
uses the advanced search option to construct a boolean query 
that takes advantage of the underlying structure of the RDF 
data in the database. She searches Diseasome for 
diseases whose class is ``Cardiovascular" or for which the associated gene is “ABCA1.” Using the QueryMed advanced 
search interface, the complex SPARQL query corresponding 
to her question is automatically constructed (Figure \ref{fig:SPARQL_adv}), and she can 
view the query results conveniently displayed in the Results Browser.

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT distinct ?disease WHERE { 
{?x <http://www.w3.org/2000/01/rdf-schema#label> 
?disease 
FILTER regex(?disease, 
"coronary artery disease", "i"). 
?x <http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseasome/class> 
<http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseaseClass/Cardiovascular>} 
UNION { ?x <http://www4.wiwiss.fu-berlin.de/
diseasome/resource/diseasome/associatedGene>
 <http://www4.wiwiss.fu-berlin.de/diseasome/
 resource/genes/ABCA1>.}
}
\end{Verbatim}
\caption{A complex SPARQL query that was dynamically constructed using the advanced query feature of the QueryMed system.}
\label{fig:SPARQL_adv}
\end{figure}

\begin{figure*}
\centering
\epsfig{file=images/add_props.jpg, width=6in}
\caption{The advanced search feature allows the user to perform exact or pattern-matching queries connected by user-specified logical operators over specific properties in given resources, taking advantage of the structure in the RDF data.}
\label{fig:add_props}
\end{figure*}

@@

\begin{figure}[h]
\centering
\epsfig{file=images/facet_creator.jpg, width=3in}
\caption{Dynamic Facet Creator and Filter}
\label{fig:facet_creator}
\end{figure}

@@

\item \textbf{Source Manager:} The Source Manager reads the source list and populates the default query list on the user interface. It also manages the default endpoints, the currently selected endpoints, and the endpoints that have been dynamically added.

@@

% The user has to specify the data source(s) to query, and can give specific values for one or many of the properties available for the data available to restrict the query space.

%\item \textbf{Additional Services:} In order to make the user interface more user friendly, the following two services perform some ``behind the scenes" work.

%\begin{enumerate}
%\item \textbf{Get Properties:} This service takes as input an individual source, and returns an array of all properties for that source.
%\item \textbf{Get Property Values:} This service takes as input an individual source and a specific property.  It returns a list of the possible values that this property can take.
%\end{enumerate}

@@

\item The proxy server, and not the Javascript client, executes SPARQL queries at the remote endpoints. This avoids cross domain XML-HTTP-Request errors in accessing web servers at various domains that are used to host the remote biomedical SPARQL endpoints.

@@

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT ?projection_1 ?projection_2 ... 
	?projection_n
WHERE {?x source:property ? projection 
	FILTER regex(? projection, '" + 
	input +"', 'i)
}
... //Other property filers are to follow
\end{Verbatim}
\caption{The structure of the SPARQL query constructed when the user selects ``Query All".}
\label{fig:min_query}
\end{figure}

@@

When the user runs an advanced query, the proxy server takes as input the data sources to be queried, the properties to be queried, the values for each of the selected properties, and the relationship between the property-value pairs and the original user-specified query term. This information is also passed to the proxy server as a JSON object.

@@

The basic query interface is designed to be as simple as possible, so that users who have little previous experience running queries on SPARQL endpoints can easily and rapidly find query results.  The advanced query functionality of our system allows users the flexibility to intuitively construct complex SPARQL queries. 
\subsubsection{Implementation}

@@

For all selected endpoints, without local data (i.e. no cache on the proxy server), the property values took longer to load.  The difference in running time between \emph{with local data} and \emph{without local data} is approximately three orders of magnitude.  The network latency and the slow performance of the remote SPARQL endpoints as compared to the local data cache explains  the increase in running time for the queries in the two cases.  We also noted that the running time of the subsequent iterations of the same query is significantly less than the initial query, probably due to browser caching. When comparing query execution times across the data sources, we see that the Drugbank takes the longest, probably due to the greater size of the Drugbank dataset as compared to the other three datasets.\footnote{As of February 15\textsuperscript{th} 2010, Diseasome contains 91,182 triples, DailyMed contains 164,276 triples, Sider contains 192,515 triples, and DrugBank contains 765,936 triples) \cite{Diseasome, DailyMed, Sider, Drugbank}}

@@

Based on the dramatic improvement in running time obtained using caching, the slowest step in executing queries with our system is retrieving triples from slow remote SPARQL endpoints.  By managing our own cache to contain the data most likely to be needed on the proxy server, we were able to decrease running time considerably.

@@

\begin{figure}
\begin{Verbatim}[frame=single]
SELECT DISTINCT ?property
WHERE { [] ?property [] }
ORDER BY ?property
\end{Verbatim}
\caption{SPARQL query to retrieve all properties. This was used to measure the performance of implementing a local cache.}
\label{fig:SPARQL_props}
\end{figure}

\begin{figure}
\centering
\epsfig{file=images/chart.jpg, width=3in}
\caption{Comparison of query execution times \emph{(ms)} with and without local data caching. The data points represent the average of three trials.}
\label{fig:chart}
\end{figure}

\pagebreak

@@

\begin{table*}
\begin{center}
\begin{tabular}{| p{1.8cm} | p{2cm} | p{1.8cm} | p{1.8cm}| p{1.8cm}| p{1.8cm}| p{1.8cm}| p{1.8cm} |}
\hline
 & \textbf{Hybrid Interface (Combines Querying \& Browsing)} & \textbf{Provides Local Caching} & \textbf{Queries Multiple Sources} & \textbf{Dynamic Addition of Sources} & \textbf{Allows Keyword Queries} & \textbf{Open Source} & \textbf{GUI} \\
\hline
\textbf{QueryMed}  & Yes & Yes & Yes & Yes & Yes & Yes & Yes  \\
\hline
\textbf{SMART}  & No & Yes & Yes & No & No & Yes & Yes  \\
\hline
\textbf{DARQ}  & No &  No & Yes & Yes & No & Yes & No  \\
\hline
\textbf{GoWeb}   & Yes & No  & Yes & No & Yes & No & Yes  \\
\hline
\textbf{BioGateway}   & Yes & Yes & Yes & No & No & Yes & Yes  \\
\hline
\textbf{Twinkle}   & No & No & Yes & No & No & Yes & Yes  \\
\hline
\end{tabular}
\caption{Comparison of selected features of the QueryMed system with other related systems.}
\label{comparison}
\end{center}
\end{table*}

A number of existing tools aim to provide a user-friendly interface for browsing biomedical semantic web data, or to allow users to perform federated queries.  Several of these are described below.

@@

The SMART query tool is a web-based application designed to allow biologists to run SPARQL queries over multiple endpoints.  Queries to the SMART system are written in the natural-language like Manchester OWL syntax \cite{Battista}.  Major differences between the QueryMed system and SMART include the ability to construct queries intuitively by specifying keywords for user selected properties, to iteratively refine query results, and to dynamically add additional endpoints in the QueryMed system.

@@

GoWeb and BioGateway are two additional systems designed for answering queries on biomedical data  \cite{Dietze, Antezana}.  GoWeb allows users to perform a hybrid search, running keyword-based queries and then filtering based on ontological concepts. However, while GoWeb functions as a search engine that incorporates ontological background knowledge to improve search results, QueryMed is a query translation system that utilizes user input in constructing SPARQL queries.  BioGateway provides a web interface to query a provided SPARQL endpoint that includes graphs from several biomedical resources.  However, BioGateway does not allow the user to run queries over multiple endpoints or to dynamically add endpoints.

@@

Most SPARQL query engines are designed to run queries against individual endpoints. But it is often useful to draw on multiple web resources in answering a query. The DARQ  system \cite{Quilitz} is designed to allow the user to run integrated queries against multiple SPARQL endpoints. But it does not offer a graphical user interface to facilitate use by biomedical domain experts who are not familiar with SPARQL query syntax.
